Evolutionary rates of SARS-CoV-2

### 2022-07-30 manuscript/variant_rates.tex
additions to ms

Here, we investigate the rate at which the SARS-CoV-2 genome changes within variants and compare this rate to the rate at which the variants themselves accumulate differences relative to reference sequence.

@@

Rates of synonymous change are very consistent across variants and also agree with the overall rate of synonymous changes of about 5-8 changes per genome per year.
Rates of non-synonymous changes are much more variable.
Within clades, the rate of non-synonymous changes varies between 5 and 16 changes per year.
Earlier clades are estimated to have larger rates around 10-15 changes per year, while rate estimates for later clade fall  between 3 to 9 changes per year.
In contrast, the int-clade non-synonymous rate exceeds 25 changes per year.

@@

\subsection*{Diversification within variants}

>>>>

Here, we investigate the rate at which the SARS-CoV-2 genome changes within variants and compare this rate to the rate at which the variants themselves accumulate differences relative to reference sequence.
This comparison reveals a clear dichotomy between slow within-variant evolution and rapid adaptive evolution giving rise to new variants.
The rate of synonymous changes between and within variants agrees, while the non-synonymous rate within variants is far below the between variant rate.
Early variants display more rapid non-synonymous evolution that later variants.

@@

Rates of synonymous change are very consistent across variants and also agree with the overall rate of synonymous changes of about 5-8 changes per genome per year.
Rates of non-synonymous changes are much more variable.
Within clades, the rate of non-synonymous changes varies between 5 and 16 changes per year.
Earlier clades are estimated to have larger rates around 10-15 changes per year, while rate estimates for later clade fall  between 3 to 9 changes per year.
In contrast, the inter-clade non-synonymous rate exceeds 25 changes per year.

@@

The apparent time dependence of non-synonymous rate suggests that at the beginning a large fraction of amino acid changes were beneficial to viral replication, not just the D614G mutation in the spike protein \citep{korber_tracking_2020}.
This applies both to the changes within clades that were ultimately displaced by more successful variants, and inter-variant changes where the fraction of adaptive changes is even higher.
Adaptive evolution complicates phylodynamic analysis, which typically assumes that the mutation process is independent of the spread and epidemiology.
This assumption is true for neutral mutations that occur along every lineage with the same rate.
Adaptive and deleterious mutations, however, affect the ability of the virus to spread.
Deleterious mutations are relatively straightforward to account for: Strongly deleterious mutations don't spread and lead to an evolutionary rate that is smaller than the mutation rates.
Weakly deleterious mutations can spread, but lineages that carry them tend to be short-lived.
A spectrum of mutations with different deleterious effects leads to time dependent effective evolutionary rates \citep{wertheim_purifying_2011}.
The intra-clade rates in Fig.~\ref{fig:rate_progression} are all estimated over a similar time frame and the time dependent rate phenomenon should not apply.

Adaptive evolution, however, is much harder to account for properly and can lead to more dramatic distortions.
Since the number of sites that allow beneficial mutations is small, adaptive evolution tends to be very stochastic.
Unlike neutral evolution, the rate of adaptive evolution depends on the population size and the transmission process.
Epistatic interactions can further steepen the dependence on population size and stochasticity.

\subsection*{Diversification within variants}
The analysis of evolutionary rates above only considers the mean divergence within rates.
